---
title: "Project 7: Difference-in-Differences and Synthetic Control"
output: html_document
---


```{r}
# Install and load packages 
if (!require("pacman")) install.packages("pacman")

devtools::install_github("ebenmichael/augsynth")

pacman::p_load(# Tidyverse packages including dplyr and ggplot2 
               tidyverse,
               ggthemes,
               augsynth,
               gsynth)
library(dplyr)

# set seed
set.seed(1)

# load data
medicaid_expansion <- read_csv('data/medicaid_expansion.csv')
medicaid_expansion
```

# Introduction

For this project, you will explore the question of whether the Affordable Care Act increased health insurance coverage (or conversely, decreased the number of people who are uninsured). The ACA was passed in March 2010, but several of its provisions were phased in over a few years. The ACA instituted the "individual mandate" which required that all Americans must carry health insurance, or else suffer a tax penalty. There are four mechanisms for how the ACA aims to reduce the uninsured population:

- Require companies with more than 50 employees to provide health insurance.
- Build state-run healthcare markets ("exchanges") for individuals to purchase health insurance.
- Provide subsidies to middle income individuals and families who do not qualify for employer based coverage.
- Expand Medicaid to require that states grant eligibility to all citizens and legal residents earning up to 138\% of the federal poverty line. The federal government would initially pay 100\% of the costs of this expansion, and over a period of 5 years the burden would shift so the federal government would pay 90\% and the states would pay 10\%.

In 2012, the Supreme Court heard the landmark case NFIB v. Sebelius, which principally challenged the constitutionality of the law under the theory that Congress could not institute an individual mandate. The Supreme Court ultimately upheld the individual mandate under Congress's taxation power, but struck down the requirement that states must expand Medicaid as impermissible subordination of the states to the federal government. Subsequently, several states refused to expand Medicaid when the program began on January 1, 2014. This refusal created the "Medicaid coverage gap" where there are indivudals who earn too much to qualify for Medicaid under the old standards, but too little to qualify for the ACA subsidies targeted at middle-income individuals.

States that refused to expand Medicaid principally cited the cost as the primary factor. Critics pointed out however, that the decision not to expand primarily broke down along partisan lines. In the years since the initial expansion, several states have opted into the program, either because of a change in the governing party, or because voters directly approved expansion via a ballot initiative.

You will explore the question of whether Medicaid expansion reduced the uninsured population in the U.S. in the 7 years since it went into effect. To address this question, you will use difference-in-differences estimation, and synthetic control.

# Data

The dataset you will work with has been assembled from a few different sources about Medicaid. The key variables are:

- **State**: Full name of state
- **Medicaid Expansion Adoption**: Date that the state adopted the Medicaid expansion, if it did so.
- **Year**: Year of observation.
- **Uninsured rate**: State uninsured rate in that year.

# Exploratory Data Analysis

Create plots and provide 1-2 sentence analyses to answer the following questions:

- Which states had the highest uninsured rates prior to 2014? The lowest?
- Which states were home to most uninsured Americans prior to 2014? How about in 2021? **Note**: 2010 state population is provided as a variable to answer this question. In an actual study you would likely use population estimates over time, but to simplify you can assume these numbers stay about the same.

```{r}
# highest and lowest uninsured rates
medicaid_expansion %>%
  filter(year < 2014) %>%
  group_by(State) %>%
  summarise(mean_uninsured = mean(uninsured_rate)) %>%
  ggplot(aes(x = reorder(State, mean_uninsured), y = mean_uninsured, fill = mean_uninsured)) +
  geom_bar(stat="identity") +
  coord_flip() +
  theme_fivethirtyeight() +
  theme(axis.title = element_text()) +
  labs(y="Mean Uninsured Rate", x = "State", fill = "Mean Uninsured Rate") +
  ggtitle("Mean Uninsured Rates before 2014")
```

The states with the highest uninsured rates prior to 2014 are Nevada, Florida and New Mexico. The lowest are Massachusetts, Hawaii and the District of Columbia. 
 

```{r}
# most uninsured Americans
medicaid_expansion %>%
  filter(year==2014 & State!= "District of Columbia") %>%
  mutate(num_uninsured = population*uninsured_rate) %>%
  ggplot(aes(x = reorder(State, num_uninsured), y = num_uninsured, fill = uninsured_rate)) +
  geom_bar(stat="identity") +
  coord_flip() +
  scale_y_continuous(labels = scales::comma) +
  theme_fivethirtyeight() +
  theme(axis.title = element_text()) +
  labs(y="Number Uninsured", x = "State", fill = "Uninsured Rate") +
  ggtitle("Number of Uninsured Americans in 2013")
  
medicaid_expansion %>%
  filter(year == 2020 & State!= "District of Columbia") %>% 
  mutate(num_uninsured = population*uninsured_rate) %>%
  ggplot(aes(x = reorder(State, num_uninsured), y = num_uninsured, fill = uninsured_rate)) +
  geom_bar(stat="identity") +
  coord_flip() +
  scale_y_continuous(labels = scales::comma) +
  theme_fivethirtyeight() +
  theme(axis.title = element_text()) +
  labs(y="Number Uninsured", x = "State", fill = "Uninsured Rate") +
  ggtitle("Number of Uninsured Americans in 2020")
```

The states with the most uninsured Americans in 2013 are Texas, California, Florida and New York. In 2020, the states with the most uninsured Americans are similar (Texas, California, Florida and Georgia) but there was a sizable reduction in the number of uninsured people in both California and Florida. In addition, New York dropped from the 4th to the 6th state with the most uninsured residents.
  
# Difference-in-Differences Estimation

## Estimate Model

Do the following:

- Choose a state that adopted the Medicaid expansion on January 1, 2014 and a state that did not. **Hint**: Do not pick Massachusetts as it passed a universal healthcare law in 2006, and also avoid picking a state that adopted the Medicaid expansion between 2014 and 2015.
- Assess the parallel trends assumption for your choices using a plot. If you are not satisfied that the assumption has been met, pick another state and try again (but detail the states you tried).

```{r}
# Parallel Trends plot

medicaid_expansion %>%
  filter(State %in% c("Maryland","Virginia")) %>%
  ggplot() + 
  geom_point(aes(x = year, 
                 y = uninsured_rate, 
                 color = State)) +
  geom_line(aes(x = year, 
                y = uninsured_rate, 
                color = State)) +
   geom_vline(aes(xintercept = 2014)) +
  theme_fivethirtyeight() +
  theme(axis.title = element_text()) +
  ggtitle('Maryland and Virginia uninsured rate \n before/after Medicaid expansion') +
  xlab('Year') +
  ylab('Uninsured Rate')
```

Initially we selected Maryland (adopted 01-01-2014) and Virginia (adopted 01-01-2019) as comparative units given their geographical proximity. The lines somewhat move together, however the pre-treatment dissimilarity between 2008-2010 are concerning. Given this, we decided to try out Utah and Colorado to ascertain if they have more aligned parallel trends.

```{r}
# Parallel Trends plot

medicaid_expansion %>%
  filter(State %in% c("Colorado","Utah")) %>%
  ggplot() + 
  geom_point(aes(x = year, 
                 y = uninsured_rate, 
                 color = State)) +
  geom_line(aes(x = year, 
                y = uninsured_rate, 
                color = State)) +
   geom_vline(aes(xintercept = 2014)) +
  theme_fivethirtyeight() +
  theme(axis.title = element_text()) +
  ggtitle('Colorado and Utah uninsured rate \n before/after Medicaid expansion') +
  xlab('Year') +
  ylab('Uninsured Rate')
```

Secondly, we selected Colorado (adopted 01-01-2014) and Utah (adopted 01-01-2020) as comparative units also given their geographical proximity to see if their trends were more parallel. There is a significant pre-treatment gap between 2008-2010; however, following 2010, the lines generally track closely together.

- Estimates a difference-in-differences estimate of the effect of the Medicaid expansion on the uninsured share of the population. You may follow the lab example where we estimate the differences in one pre-treatment and one post-treatment period, or take an average of the pre-treatment and post-treatment outcomes

```{r}
# Difference-in-Differences estimation
# Colorado-Utah
medi <- medicaid_expansion %>%
  filter(State %in% c("Colorado","Utah")) %>%
  filter(year >= 2013 & year<= 2015) 

# pre-treatment difference

pre_diff <- medi %>%
  filter(year == 2013) %>%
  select(State, 
         uninsured_rate) %>%
  spread(State, 
         uninsured_rate) %>%
  summarise(Utah - Colorado)
  
# post-treatment difference

post_diff <- medi %>%
  filter(year == 2015) %>%
  select(State, 
         uninsured_rate) %>%
  spread(State, 
         uninsured_rate) %>%
  summarise(Utah - Colorado)

# diff-in-diffs

diff_in_diffs <- post_diff - pre_diff
diff_in_diffs

```

## Discussion Questions

- Card/Krueger's original piece utilized the fact that towns on either side of the Delaware river are likely to be quite similar to one another in terms of demographics, economics, etc. Why is that intuition harder to replicate with this data?
- **Answer**: That intuition is harder with this data because of the geographic clustering in the passage of Medicaid expansion. It is more likely that neighboring states have the same treatment status. For the state of Oregon, for example, it may make most sense to compare it to the state of Washington, but Washington also passed Medicaid expansion. In a similar vein, most Southern states did not pass the Medicaid expansion so it is impossible to compare the outcomes for Mississippi and Alabama, for example, because both did not pass an expansion. There are also likely important observable and unobservable differences in states that passed (e.g., those in New England and the West) and those that did not a Medicaid expansion (e.g., a handful of Midwestern and Southern states). There are also several different states that passed Medicaid expansions (rather than a single city or state) so it is not practical or possible to find an appropriate "control" for each treated state.

- What are the strengths and weaknesses of using the parallel trends assumption in difference-in-differences estimates?
- **Answer**: While straighforward to understand and implement, the parallel trends assumption is not a testable hypothesis.

# Synthetic Control

Estimate Synthetic Control

Although several states did not expand Medicaid on January 1, 2014, many did later on. In some cases, a Democratic governor was elected and pushed for a state budget that included the Medicaid expansion, whereas in others voters approved expansion via a ballot initiative. The 2018 election was a watershed moment where several Republican-leaning states elected Democratic governors and approved Medicaid expansion. In cases with a ballot initiative, the state legislature and governor still must implement the results via legislation. For instance, Idaho voters approved a Medicaid expansion in the 2018 election, but it was not implemented in the state budget until late 2019, with enrollment beginning in 2020.

Do the following:

- Choose a state that adopted the Medicaid expansion after January 1, 2014. Construct a non-augmented synthetic control and plot the results (both pre-treatment fit and post-treatment differences). Also report the average ATT and L2 imbalance.

**For this analysis**, we chose California, as it adapted the Medicaid expansion on January 1st, 2014. 

```{r}
# non-augmented synthetic control
# drop states that adopted medicaid before January 1st 2014
medicaid_expansion <- medicaid_expansion %>% 
  mutate(year_adopted = format(Date_Adopted, format="%Y"))
         
medi2<- medicaid_expansion %>%
  filter(State=="California" | is.na(Date_Adopted))

# Create a treatment indicator for California after January 1 2014
medi3 <- medi2 %>%
  mutate(treatment = ifelse(State == "California" & year >= 2014, 1, 0))


#run sysnthetic control
syn <- augsynth(uninsured_rate ~ treatment, State, year, medi3,
                progfunc = "None", scm = T)

summary(syn)
plot(syn)

# Explore states that contributed weights weights
data.frame(syn$weights) %>%
  # change index to a column
  tibble::rownames_to_column('State') %>%
  ggplot() +
  geom_bar(aes(x = State, 
               y = syn.weights),
           stat = 'identity') +
  theme_fivethirtyeight() +
  theme(axis.title = element_text(),
        axis.text.x = element_text(angle = 90)) +
  ggtitle('Synthetic Control Weights') +
  xlab('State') +
  ylab('Weight')

```

We now plot the pre- and post-treatment differences

```{r}
# extract the data from the output
syn_sum <- summary(syn)

syn_ca <- medi3 %>%
  filter(State == "California") %>%
  bind_cols(difference = syn_sum$att$Estimate) %>%
  mutate(synthetic_california = uninsured_rate + difference)

# Plot
syn_ca %>%
  ggplot() +
  geom_point(aes(x = year, 
                 y = uninsured_rate, 
                 color = 'California')) +
  geom_line(aes(x = year, 
                y = uninsured_rate, 
                color = 'California')) +
  geom_point(aes(x = year, 
                 y = synthetic_california, 
                 color = 'Synthetic California')) +
  geom_line(aes(x = year, 
                y = synthetic_california, 
                color = 'Synthetic California')) +
  scale_color_manual(values = c('California' = 'red', 'Synthetic California' = 'blue')) +
  geom_vline(aes(xintercept = 2014)) +
  theme_fivethirtyeight() +
  theme(axis.title = element_text()) +
  ggtitle('California and Synthetic California') +
  xlab('Year') +
  ylab('Uninsured Rate')
```


**Answer**: The Average ATT estimate was -0.050, with an L2 imbalance of 0.003.

- Re-run the same analysis but this time use an augmentation (default choices are Ridge, Matrix Completion, and GSynth). Create the same plot and report the average ATT and L2 imbalance.

```{r}
# augmented synthetic control
ridge_syn <- augsynth(uninsured_rate ~ treatment, State, year, medi3,
                progfunc = "ridge", scm = T)

summary(ridge_syn)
plot(ridge_syn)

```

- Plot barplots to visualize the weights of the donors.

```{r}
# barplots of weights
data.frame(ridge_syn$weights) %>%
  tibble::rownames_to_column('State') %>%
  ggplot() +
  geom_bar(aes(x = State, 
               y = ridge_syn.weights),
           stat = 'identity') +
  theme_fivethirtyeight() +
  theme(axis.title = element_text(),
        axis.text.x = element_text(angle = 90)) +
  ggtitle('Ridge Synthetic Control Weights') +
  xlab('State') +
  ylab('Weight') 

```

**Answer**: In the augmented synthetic control, the Average ATT estimate was -0.051, with an L2 imbalance of 0.003. 


**HINT**: Is there any preprocessing you need to do before you allow the program to automatically find weights for donor states? We needed to remove other states that passed Medicaid expansion because they cannot serve as "controls" or a California counterfactual in the absence of a Medicaid expansion if they too passed the expansion.

## Discussion Questions


- What are the advantages and disadvantages of synthetic control compared to difference-in-differences estimators?
- **Answer**: Synthetic control provides a more rigorous, less ad-hoc way of selecting control units from a large pool of potential controls than the diff-in-diff estimators. It also leverages this large pool of potential controls to conduct permutation-based inference in a manner that is robust to the possibility of unit-by-time period specific shocks. In other words, it accounts for the fact that, even if we observed the entire population for each state, there would still be some deviation between the treated unit and its synthetic control because there are aggregate (i.e., unit-level) shocks that occur at the unit-by-time level. It also allows for the estimation of standard errors.

- One of the benefits of synthetic control is that the weights are bounded between [0,1] and the weights must sum to 1. Augmentation might relax this assumption by allowing for negative weights. Does this create an interpretation problem, and how should we balance this consideration against the improvements augmentation offers in terms of imbalance in the pre-treatment period?
- **Answer**: It does not appear that any of our donor pool states had negative weights even with augmentation in our case. However, this may become an issue with interpretation because it is not intuitive what negativing weighting potential controls means. But we should balance this consideration with the improvement augmentation offers in terms of balance and parallel trends. If the treated unit happens to be a unit with an exceptionally high (or low) uninsured rate, for example, the only control units that can receive positive weights are those with very high (or low) levels of uninsured. This is undesirable in that some control units may be better at matching the trends in the treated unit, but they can't receive positive weight unless they also have high levels of the outcome. From a DiD perspective, we may be discarding units that are good matches for satisfying the parallel trends assumption simply because they aren't good matches in terms of their baseline levels of uninsured.

# Staggered Adoption Synthetic Control

## Estimate Multisynth

Do the following:

- Estimate a multisynth model that treats each state individually. Choose a fraction of states that you can fit on a plot and examine their treatment effects.

```{r}
# multisynth model states

# randomly selecting 11 states who passed Medicaid expansion (so we have 12 final plots)
`%notin%` <- Negate(`%in%`)
states2014 <- medicaid_expansion %>%
  subset(year=="2020" & !is.na(year_adopted) & State %notin% c("Massachusetts", "District of Columbia")) %>%
  select(State)

states2014 <- sample_n(states2014, 11)
states2014 <- dplyr::pull(states2014, State)

# first creating treatment indicator
medicaid_expansion_clean <- medicaid_expansion %>%
  filter(State %in% states2014 | is.na(Date_Adopted)) %>%
  mutate(expan = ifelse(year<year_adopted,0,1),
         expan = ifelse(is.na(year_adopted), 0, expan)) %>%
  arrange(State)

# with default nu (e.g., not prespecifying nu)
ppool_syn <- multisynth(uninsured_rate ~ expan, State, year, 
                        medicaid_expansion_clean, n_leads = 6)

print(ppool_syn$nu)

ppool_syn

# getting ATT balance statistics
ppool_syn_summ <- summary(ppool_syn)

# plotting
ppool_syn_summ$att %>%
  ggplot(aes(x = Time, y = Estimate, color = Level)) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = 0) +
  theme_fivethirtyeight() +
  theme(axis.title = element_text(),
        legend.position = "bottom") +
  ggtitle('Synthetic Controls for Medicaid Expansion') +
  xlab('Time') +
  ylab('Percent Uninsured') +
  theme(legend.position = "none") +
  facet_wrap(~Level)
```

- Estimate a multisynth model using time cohorts. For the purpose of this exercise, you can simplify the treatment time so that states that adopted Medicaid expansion within the same year (i.e. all states that adopted expansion in 2016) count for the same cohort. Plot the treatment effects for these time cohorts.

```{r}
# multisynth model time cohorts
# bringing back in full sample minus DC and Massachusetts
medicaid_expansion_clean <- medicaid_expansion %>%
  subset(State %notin% c("Massachusetts", "District of Columbia"))

ppool_syn_time <- multisynth(uninsured_rate ~ expan, State, year, 
                        medicaid_expansion_clean, n_leads = 6, time_cohort = TRUE)

ppool_syn_time_summ <- summary(ppool_syn_time)

ppool_syn_time_summ

ppool_syn_time_summ$att %>%
  ggplot(aes(x = Time, y = Estimate, color = Level)) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = 0) +
  theme_fivethirtyeight() +
  theme(axis.title = element_text(),
        legend.position = 'None') +
  ggtitle('Synthetic Controls for Medicaid Expansion') +
  xlab('Time') +
  ylab('Percent Uninsured') +
  facet_wrap(~Level)
```

## Discussion Questions

- One feature of Medicaid is that it is jointly administered by the federal government and the states, and states have some flexibility in how they implement Medicaid. For example, during the Trump administration, several states applied for waivers where they could add work requirements to the eligibility standards (i.e. an individual needed to work for 80 hours/month to qualify for Medicaid). Given these differences, do you see evidence for the idea that different states had different treatment effect sizes?
- **Answer**: Yes. While most states saw a decline in the share of uninsured residents in the year directly following expansion, it did not continue to decline uniformily across states. Some states even saw seemingly no effect or an increase in the uninsured rate after Medicaid expansion.


- Do you see evidence for the idea that early adopters of Medicaid expansion enjoyed a larger decrease in the uninsured population?
- **Answer**: Not really. For most states, it appears that there is a sizable drop in the uninsured rate in the year directly following Medicaid expansion but then trends diverge. Among those who passed the law first in 2014, the uninsured rate declined before ticking back up and plateauing (which is also the trendline for the average given that over two dozen states were early adopters). For the three states that passed the law in 2015, the uninsured rate appears to have declined steadily for four years following expansion. 

# General Discussion Questions

- Why are DiD and synthetic control estimates well suited to studies of aggregated units like cities, states, countries, etc?
- **Answer**: Because in applied policy settings, we are often interested in the impact of certain laws at the level of the city or the state rather than a particular individual. The counterfactual of interest is what would have happened in a given aggregated unit in the absence of the policy rather than an outcome for a specific individual in that unit. They also do not require longitudinal data at the individual-level, which is typically much harder to gather than repeated cross-sections.

- What role does selection into treatment play in DiD/synthetic control versus regression discontinuity? When would we want to use either method?
- **Answer**: Selection into treatment concerns are assuaged in DiD/synthetic control models by the parallel trends assumption. This assumption implies that any factor that is specific to a given state in the present example but does not change over time, or changes over time but changes in equal amount for both states, is netted out in the diffs-in-diffs estimator. With regression discontinuity designs, the key assumptions are (1) monotonicity (which rules out defiers) and (2) the continuity assumption (which more directly deals with selection into treatment). If we find evidence of visual breaks in the running variable or outcomes at the RD threshold, this would be cause for concern. It may be inappropriate to then assume that units right on the other side of the threshold are good "controls". Typically, RD designs rely on individual level data while DiD/syntheic controls are more suited to aggregated units like states.
